Optimal synaptic strategies for different timescales of memory

### 2016-03-11 Paper/SynapsePaper.tex
thrown in supp

%% OPTIONAL MACRO DEFINITIONS
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
% indices
%
% brackets etc.
\newcommand{\prn}[1]{\left ( #1 \right )}
\newcommand{\brc}[1]{\left\{ #1 \right\}}
\newcommand{\brk}[1]{\left [ #1 \right ]}
\newcommand{\abs}[1]{\left\lvert #1 \right\rvert}
\newcommand{\nrm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\avc}[2]{\left\langle #1 \,\middle | \, #2 \right\rangle}
%\newenvironment{cases}{\left\{\begin{aligned}}{\end{aligned}\right.}
%
% Sets
\newcommand{\set}[2]{\left\{ #1 \,\middle |\, #2 \right\}}
\newcommand{\cond}[2]{\!\left( #1 \,\middle | \, #2 \right)}
\newcommand{\condb}[2]{\!\left[ #1 \,\middle | \, #2 \right]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][\rule{0mm}{0mm}]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][\rule{0mm}{0mm}]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][\rule{0mm}{0mm}]{\left (\frac{\partial #2}{\partial {#3}}\right )_{\!\!#1}}
\newcommand{\intd}[2][\rule{0mm}{0mm}]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
% Referencing sections, figures, etc
\newcommand{\sref}[1]{\S\ref{#1}}
\newcommand{\fref}[1]{fig.\ref{#1}}
\newcommand{\Fref}[1]{Fig.~\ref{#1}}
\newcommand{\tref}[1]{tab.\ref{#1}}
\newcommand{\Tref}[1]{Tab.\ref{#1}}
%
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

%\renewcommand\Fref[1]{}
\providecommand\hyperref[2][1]{#2}

@@

>>>>

%% OPTIONAL MACRO DEFINITIONS
\newcommand{\ra}{\rightarrow}
\newcommand{\lr}{\leftrightarrow}
\newcommand{\cdt}{\!\cdot\!}
\newcommand{\vp}{\vspace{0.5cm}}
\newcommand{\hp}{\hspace{0.5cm}}
%
%e.g., i.e. with normal spaces
\newcommand{\eg}{e.g.\ }
\newcommand{\ie}{i.e.\ }
\newcommand{\cf}{cf.\ }
\newcommand{\etc}{etc.\ }
\newcommand{\wrt}{wrt.\ }
\newcommand{\Eg}{E.g.\ }
\newcommand{\Ie}{I.e.\ }
\newcommand{\Cf}{Cf.\ }
\newcommand{\Etc}{Etc.\ }
%
% indices
%
% brackets etc.
\newcommand{\prn}[1]{\left ( #1 \right )}
\newcommand{\brc}[1]{\left\{ #1 \right\}}
\newcommand{\brk}[1]{\left [ #1 \right ]}
\newcommand{\abs}[1]{\left\lvert #1 \right\rvert}
\newcommand{\nrm}[1]{\left\lVert #1 \right\rVert}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\avc}[2]{\left\langle #1 \,\middle | \, #2 \right\rangle}
%\newenvironment{cases}{\left\{\begin{aligned}}{\end{aligned}\right.}
%
% Sets
\newcommand{\set}[2]{\left\{ #1 \,\middle |\, #2 \right\}}
\newcommand{\cond}[2]{\!\left( #1 \,\middle | \, #2 \right)}
\newcommand{\condb}[2]{\!\left[ #1 \,\middle | \, #2 \right]}
%
% Derivatives, etc. First argument is optional.
\newcommand{\diff}[3][\rule{0mm}{0mm}]{\frac{\mathrm{d}^{#1} #2}{\mathrm{d}{#3}^{#1}}}
\newcommand{\pdiff}[3][\rule{0mm}{0mm}]{\frac{\partial^{#1} #2}{\partial {#3}^{#1}}}
\newcommand{\pdiffc}[3][\rule{0mm}{0mm}]{\left (\frac{\partial #2}{\partial {#3}}\right )_{\!\!#1}}
\newcommand{\intd}[2][\rule{0mm}{0mm}]{\int #1\!\!\dr #2\,}
%
% Un-italicised letters
\newcommand{\dr}{\mathrm{d}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\ir}{\mathrm{i}}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\Tr}{Tr}
\DeclareMathOperator{\Det}{Det}
\DeclareMathOperator{\cov}{Cov}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\bias}{Bias}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\sech}{sech}
\DeclareMathOperator{\diag}{diag}
%
% The default \Im and \Re look crap
\renewcommand{\Im}{\operatorname{\mathfrak{Im}}}
\renewcommand{\Re}{\operatorname{\mathfrak{Re}}}
%
% logic
\newcommand{\means}{\Longleftrightarrow}
\newcommand{\requires}{\Longleftarrow}
%
% Referencing sections, figures, etc
\newcommand{\sref}[1]{\S\ref{#1}}
\newcommand{\fref}[1]{fig.\ref{#1}}
\newcommand{\Fref}[1]{Fig.~\ref{#1}}
\newcommand{\tref}[1]{tab.\ref{#1}}
\newcommand{\Tref}[1]{Tab.\ref{#1}}
%
%\input{sl_symbols.tex}
%
%additional symbols:
\newcommand{\half}{\frac{1}{2}}
\newcommand{\CO}{\mathcal{O}}
\newcommand{\CL}{\mathcal{L}}
\newcommand{\dt}{\dr t}
%
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\newcommand{\wv}{\vec{w}}
\newcommand{\wvi}{\vec{w}_\text{ideal}}
%matrices
\newcommand{\inv}{^{-1}}
\newcommand{\dg}{^\mathrm{dg}}
\newcommand{\trans}{^\mathrm{T}}
\newcommand{\I}{\mathbf{I}}
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distributions
\newcommand{\prob}{\mathbf{p}}
\newcommand{\eq}{\prob^\infty}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols for matrices
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{q}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%sets
\newcommand{\CS}{\mathcal{S}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
\newcommand{\comp}{^\mathrm{c}}
%eigenmodes
\newcommand{\uv}{\mathbf{u}}
\newcommand{\vv}{\mathbf{v}}
\newcommand{\CI}{\mathcal{I}}

% ----------------------------------------------------------------
\input{sl_theorems_preamble.tex}
% ----------------------------------------------------------------
%\renewcommand\Fref[1]{}
\providecommand\hyperref[2][1]{#2}
\providecommand\texorpdfstring[2]{#1}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\begin{center}
  \textbf{\Large Supplementary material}
\end{center}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Here we provide more details underlying the derivations of results in the main paper.

\section{Continuous time Markov processes}\label{sec:ContMarkov}

In this section we'll provide a summary of all the relevant properties of ergodic Markov chains in continuous time to define notation.
It is a generalization of material that can be found in \cite{kemeny1960finite} with some ideas from \cite{hunter2000survey}.

\subsection{Notation}\label{sec:not}

For any matrix $\mathbf{A}$, we define matrices $\mathbf{A}\dg$ and $\overline{\mathbf{A}}$ as
%
\begin{equation}\label{eq:dgdef}
  \mathbf{A}\dg_{ij} \equiv \delta_{ij}\mathbf{A}_{ij},
  \qquad
  \overline{\mathbf{A}} \equiv \mathbf{A}-\mathbf{A}\dg.
\end{equation}
%
We let $\onev$ denote a column-vector of ones and $\onem=\onev\onev\trans$ denote a matrix of ones.

A continuous time Markov process is described by a matrix of transitions rates, $\MM_{ij}$, from state $i$ to $j$ with row sums equal to zero ($\MM\onev=0$).
The probabilities of being in each state at time $t$, the row-vector $\mathbf{p}(t)$, evolve according to
%
\begin{equation}\label{eq:rowsum}
  \diff{\mathbf{p}(t)}{t} = \mathbf{p}(t) \MM,
\end{equation}
%
where $\mathbf{p}(t)\onev=1$.

The equilibrium probabilities, $\eq$, satisfy
%
\begin{equation}\label{eq:equilibrium}
  \eq\MM=0,
  \qquad
  \eq\onev=1.
\end{equation}
%
As we assume an ergodic process, this eigenvalue is non-degenerate.
If all other eigenvalues have strictly negative real parts, the process is regular (aperiodic).

We define additional matrices
%
\begin{equation}\label{eq:defDLP}
  \Lb \equiv (-\MM\dg)\inv,
  \qquad
  \Pb \equiv \I + \Lb\MM.
\end{equation}
%
It can be shown that $\Lb_{ii}$ is the mean time it takes to leave state $i$ and $\Pb_{ij}$ is the probability the the next transition from state $i$ goes to state $j$:
%
\begin{equation}\label{eq:LamdaPcmpt}
  \Lb_{ii} = \frac{1}{\sum_{j \neq j} \MM_{ij}},
  \qquad
  \Pb_{ij} = %\left\{
%  \begin{aligned}
%     &0                                         && \text{if }i=j, \\
%     &\frac{\MM_{ij}}{\sum_{k \neq j} \MM_{ik}} &&\text{otherwise}.
%  \end{aligned}
%  \right.
  \begin{cases}
     0                                         &\text{if }i=j, \\
     \frac{\MM_{ij}}{\sum_{k \neq j} \MM_{ik}} &\text{otherwise}.
  \end{cases}
\end{equation}
%
Furthermore, we also define
%
\begin{equation}\label{eq:pdotD}
  \D \equiv \operatorname{diag}(\eq)\inv,
  \qquad\implies\qquad
  \eq\D=\onev\trans.
\end{equation}
%

\subsection{Fundamental matrix}\label{sec:fund}

For our results below regarding the integral of the memory curve, it can be useful to invert the stochastic transition matrix, $\MM$.
However, since $\MM$ has a zero eigenvalue, it cannot be inverted.
For this reason, the fundamental matrix arises as a useful surrogate for the inverse of $\MM$.
It is related to the first passage times, as we will see in the \hyperref[sec:fpt]{next subsection}.
Here we define the fundamental matrix and review its properties.

\begin{defn}[Fundamental matrix]
  For discrete time, the generalized fundamental matrix was defined in \cite{Kemeny1981fund}.
  For continuous time, we define:
  %
  \begin{equation}\label{eq:funddef}
    \fund \equiv (-\MM + \onev\pib)\inv,
  \end{equation}
  %
  where $\pib$ is any row-vector with $\pib\onev=1/\tau\neq0$.
\end{defn}
Note that the canonical choice for the discrete time version, $\pib=\eq$, is not available here due to problems with units.
It will be helpful to choose $\pib$ to be independent of $\MM$, \eg $\pib=\onev\trans/(n\tau)$.
All quantities that we calculate using $\fund$ below will be independent of this choice.

\begin{thm}
  The definition of $\fund$ is valid, \ie $(-\MM + \onev\pib)$ is invertible.
\end{thm}
\begin{proof}
  Assume there exists an $\mathbf{x}$ such that
  %
  \begin{equation}\label{eq:fundinvkern}
    (-\MM + \onev\pib)\mathbf{x}=0.
  \end{equation}
  %
  Multiplying from the left with $\eq$ gives
  %
  \begin{equation}\label{eq:fundinvpix}
    \pib \mathbf{x} = 0.
  \end{equation}
  %
  Substituting back into \eqref{eq:fundinvkern} gives
  %
  \begin{equation*}
    \MM\mathbf{x}=0.
  \end{equation*}
  %
  As we assume an ergodic process, the zero eigenvalue is non-degenerate.
  Therefore, $\mathbf{x}=\lambda\onev$.
  Substituting this into \eqref{eq:fundinvpix} gives
  %
  \begin{equation*}
    \lambda\pib\onev = \frac{\lambda}{\tau} = 0.
  \end{equation*}
  %
  As we defined $\pib$ such that $1/\tau\neq0$, this means $\lambda=0 \implies \mathbf{x}=0$.
\end{proof}

\begin{cor}
  %
  \begin{align}
    \pib\fund &= \eq, \label{eq:fundprob}\\
    \fund\onev &= \tau\onev,\label{eq:fundrowsum}\\
    \I+\MM\fund &= \onev\eq, \label{eq:fundQZ}\\
    \I+\fund\MM &= \tau\onev\pib. \label{eq:fundZQ}
  \end{align}
  %
\end{cor}
\begin{proof}
  We can deduce \eqref{eq:fundprob} and \eqref{eq:fundrowsum} be pre/post-multiplying the following equations by $\fund$:
  %
  \begin{equation*}
    \begin{aligned}
      \eq(-\MM + \onev\pib) &= \pib, \\
      (-\MM + \onev\pib)\onev &= \frac{\onev}{\tau}.
    \end{aligned}
  \end{equation*}
  %
  We can then deduce \eqref{eq:fundQZ} and \eqref{eq:fundZQ} by substituting these into
  %
  \begin{equation*}
    (-\MM + \onev\pib)\fund = \fund(-\MM + \onev\pib) = \I.
  \end{equation*}
  %
\end{proof}

%We will not need to use the concept of a g-inverse anywhere, so you can skip ahead to \sref{sec:fpt} now. For the sake of completeness, we can define $\MM^\#$ as
%%
%\begin{equation}\label{eq:ginvdef}
%  \MM^\# \equiv \tau\onev\eq - \fund
%\end{equation}
%%
%\begin{thm}
%  The matrix $\MM^\#$ satisfies the following defining properties of a g-inverse of $\MM$:
%  %
%  \begin{align}
%    \MM^\# \MM \MM^\# &= \MM^\#,     \label{eq:ginvpropo} \\
%    \MM \MM^\# \MM    &= \MM,        \label{eq:ginvpropi} \\
%    \MM \MM^\#        &= \MM^\# \MM, \label{eq:ginvcomm}
%  \end{align}
%  %
%  with \eqref{eq:ginvcomm} holding iff $\eq$ and $\pib$ are proportional to each other.
%\end{thm}
%\begin{proof}
%  Using \eqref{eq:rowsum}, \eqref{eq:equilibrium}, \eqref{eq:fundprob} and \eqref{eq:fundrowsum}, one can verify
%  %
%  \begin{alignat*}{2}
%    \MM \MM^\# &= -\MM \fund,
%    &\qquad
%    \pib \MM^\# &= 0,
%    \\
%    \MM^\# \MM &= -\fund \MM,
%    &
%    \MM^\# \onev &= 0.
%  \end{alignat*}
%  %
%  Substituting into \eqref{eq:fundQZ} and \eqref{eq:fundZQ} leads to
%  %
%  \begin{equation*}
%    \I - \MM \MM^\# = \onev \eq,
%    \qquad
%    \I - \MM^\# \MM = \tau \onev \pib.
%  \end{equation*}
%  %
%  Pre/post-multiplying these by $\MM$ or $\MM^\#$ produces \eqref{eq:ginvpropo} and \eqref{eq:ginvpropi}. Subtracting produces
%  %
%  \begin{equation*}
%    [\MM,\MM^\#] = \onev(\tau\pib-\eq),
%  \end{equation*}
%  %
%  which vanishes iff $\eq$ and $\pib$ are proportional to each other (due to $\pib\onev=1/\tau$ and $\eq\onev=1$, the constant of proportionality could only be $\tau$).
%\end{proof}

\subsection{First passage times}\label{sec:fpt}

\begin{defn}[First passage time matrix]
 We define $\fptb_{ij}$ as the mean time it takes the process to reach state $j$ for the first time, starting from state $i$.
 We also define $\fpt\dg_{ii}$ as the mean time it takes the process to return to state $i$.
 As usual, $\fpt=\fptb+\fpt\dg$.
\end{defn}

This matrix is given by
%
\begin{equation}\label{eq:fpt}
  \fpt=(\onem\fund\dg - \fund + \Lb)\D,
\end{equation}
%
see \cite{Yao1985fpt} for a proof.
We can separate this into its diagonal and off-diagonal pieces.

The recurrence times are given by
%
\begin{equation}\label{eq:recurtime}
  \fpt\dg = \Lb\D.
\end{equation}
%
or in component form
%
\begin{equation*}
  \eq_i \Lb_{ii}\inv \fpt\dg_{ii} = 1.
\end{equation*}
%
The extra factor of $\Lb_{ii}$, compared to the discrete case \cite[Th.4.4.5]{kemeny1960finite}, occurs because in this case we are demanding that the process leaves the initial state once before returning, whereas in the discrete case we only measure the time it takes to go to the initial state after the first time-step.

The off-diagonal mean first passage times are given by
%
\begin{equation}\label{eq:fptfund}
  \fptb = (\onem\fund\dg - \fund)\D.
\end{equation}
%
or in component form:
%
\begin{equation}\label{eq:fptfundcmpt}
  \fptb_{ij} = \frac{\fund_{jj}-\fund_{ij}}{\eq_j}.
\end{equation}
%

\subsection{Mixing time (Kemeny's constant)}\label{sec:mixtime}

\begin{thm} \mbox{}\label{th:kemenyconst}
  The quantity
  %
  \begin{equation}\label{eq:mixdef}
    \eta \equiv \sum_j \fptb_{ij}\eq_j
  \end{equation}
  %
  is independent of $i$.
\end{thm}
\begin{proof}
  For discrete time, a proof can be found in \cite[Th.4.4.10]{kemeny1960finite}.
  For continuous time, we use \eqref{eq:fptfund}, \eqref{eq:fundrowsum} and the transpose of \eqref{eq:pdotD}:
  %
  \begin{equation*}
    \begin{aligned}
      \fptb (\eq)\trans &= (\onem\fund\dg - \fund) \D (\eq)\trans \\
        &= (\onev\onev\trans\fund\dg - \fund) \onev \\
        &= (\onev\trans\fund\dg\onev)\onev - \fund\onev \\
        &= (\tr\fund - \tau)\onev.
    \end{aligned}
  \end{equation*}
  %
  which proves \eqref{eq:mixdef} with $\eta = \tr\fund - \tau$.
\end{proof}

Note that it is essential that we use $\fptb$ and not $\fpt$ here, as that would lead to $\eta_i=\eta+\Lb_{ii}$, unlike the discrete time version, where this would only shift $\eta$ by 1.

\subsection{Sensitivity of equilibrium distribution}\label{sec:sensitivity}

Suppose that the Markov process, defined by $\MM$, depends on some parameter $\alpha$.
Differentiating \eqref{eq:funddef} gives
%
\begin{equation}\label{eq:diffZ}
  \diff{\fund}{\alpha} = \fund \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can substitute this into the derivative of \eqref{eq:fundprob}:
%
\begin{equation}\label{eq:diffp}
  \diff{\eq}{\alpha} = \pib \fund \diff{\MM}{\alpha} \fund = \eq \diff{\MM}{\alpha} \fund.
\end{equation}
%
We can rewrite this in component form and use the fact that $\MM_{ii} = - \sum_{i\neq j} \MM_{ij}$:
%
\begin{equation}\label{eq:diffpT}
\begin{aligned}
  \diff{\eq_k}{\alpha} &= \sum_{i,j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} \fund_{jk} + \sum_i \eq_i \diff{\MM_{ii}}{\alpha} \fund_{ik} \\
    &= \sum_{i\neq j} \eq_i \diff{\MM_{ij}}{\alpha} (\fund_{jk} - \fund_{ik}) \\
    &= \sum_{i\neq j} \diff{\MM_{ij}}{\alpha} \eq_i \eq_k (\fptb_{ik} - \fptb_{jk}).
\end{aligned}
\end{equation}
%
This is a generalization of a result of \cite{cho2000markov} from discrete to continuous time that we will need below.
Note that the summand vanishes for $i=j$, so we can drop the restriction $i\neq j$ from the range of the sum.

\subsection{Subsets and flux}\label{sec:subsets}

Let us denote the set of states by $\CS$.
Consider a subset $\CA\subset\CS$.
We can define a projection operator onto this subset:
%
\begin{equation}\label{eq:proj}
  \prn{\I^\CA}_{ij} =
    \begin{cases}
      1 &\text{if $i=j\in\CA$,}\\
      0 &\text{otherwise.}
    \end{cases}
\end{equation}
%
We will use superscripts/subscripts to denote projection onto/summation over a subset:
%
\begin{equation}\label{eq:projsum}
  \begin{aligned}
    \pib^\CA &=\pib\I^\CA, &
    \M^{\cdot\CA}&=\M\I^\CA, &
    \M^{\CA\cdot}&=\I^\CA\M, &
    \mathbf{x}^\CA &= \I^\CA\mathbf{x},
    \\
    \pib_\CA &=\pib\onev^\CA, &
    \M_{\cdot\CA}&=\M\onev^\CA, &
    \M_{\CA\cdot}&=\prn{\onev^\CA}\trans\M, &
    \mathbf{x}_\CA &= \prn{\onev^\CA}\trans\mathbf{x},
  \end{aligned}
\end{equation}
%
where $\pib$ is a row vector, $\M$ is a matrix and $\mathbf{x}$ is a column vector.

We can define a flux matrix, a.k.a.\ ergodic flow:
%
\begin{equation}\label{eq:flux}
  \F = \D\inv\MM,
  \qquad
  \F_{ij} = \eq_i \MM_{ij}.
\end{equation}
%
This measures the flow of probability between states in the equilibrium distribution.
Detailed balance, a.k.a.\ reversibility, is equivalent to $\F=\F\trans$.

The flux between two subsets is a particularly useful quantity:
%
\begin{equation}\label{eq:subflux}
  \F_{\CA\CB} = {\eq}^\CA\MM\onev^\CB.
\end{equation}
%
One can show that
%
\begin{equation}\label{eq:compflux}
  \F_{\CA\CA\comp} = \F_{\CA\comp\CA} = -\F_{\CA\CA} = -\F_{\CA\comp\CA\comp}
\end{equation}
%
using $\prn{{\eq}^\CA + {\eq}^{\CA\comp}}\MM=0$ and $\MM\prn{{\onev}^\CA + {\onev}^{\CA\comp}}=0$.

\subsection{Lumpability}\label{sec:lump}

Suppose we have partitioned the states into disjoint subsets, $\brc{\CA_\alpha}$:
%
\begin{equation}\label{eq:partition}
  \bigcup_\alpha \CA_\alpha = \CS,
  \qquad
  \CA_\alpha \cap \CA_\beta = \delta_{\alpha\beta}\CA_\alpha.
\end{equation}
%
We will use $\alpha$ instead of $\CA_\alpha$ in superscripts and subscripts in the following.
The fact that these subsets are disjoint and exhaustive allows us to define the function
%
\begin{equation}\label{eq:whichsey}
  \sigma(i)=\alpha
  \qquad\means\qquad
  i\in\CA_\alpha.
\end{equation}
%

We can use this partition to define a new stochastic process associated with the original Markov chain.
At time $t$, if the state of the original process is $i$, the state of the new process is $\sigma(i)$.

One may ask if this new process is a Markov chain.
The answer is yes, if the original Markov chain has a property called lumpability \wrt the partition
(see \cite[\S6.3]{kemeny1960finite} for the discrete time version and \cite{burke1958markovian,Ball1993Lumpability} for continuous time):
%
\begin{equation}\label{eq:lump}
  \sigma(i)=\sigma(j)
  \quad\implies \quad
  \MM_{i\alpha}=\MM_{j\alpha} \equiv \sum_{k\in\CA_\alpha} \MM_{jk},
\end{equation}
%
\ie the total transition rate from some state to some subset is the same for all starting states within the same subset.
This common value is the transition rate for the new lumped Markov chain.

This can be rewritten with the aid of two matrices
%
\begin{equation}\label{eq:lumpmats}
  U_{\alpha i} = \frac{\delta_{\alpha\sigma(i)}}{\abs{\CA_\alpha}},
  \qquad
  V_{i\alpha} = \delta_{\sigma(i)\alpha}.
\end{equation}
%
Left multiplication by $U$ averages over subsets, right multiplication by $V$ sums over subsets.
For $U$, we chose the uniform measure in each subset. Any measure would work equally well, \eg one proportional to the equilibrium distribution:
%
\begin{equation}\label{eq:altlumpmats}
  U_{\alpha i} = \frac{{\eq_i}^\alpha}{\eq_\alpha}.
\end{equation}
%

One can show that $(UV)_{\alpha\beta}=\delta_{\alpha\beta}$.
The matrix $VU$ is also interesting.
It has a block diagonal structure, with each block corresponding to a subset.
Each block is a discrete-time ergodic Markov matrix (it is an independent trials process with probabilities given by the measure chosen for $U$).
This means that the right eigenvectors with eigenvalue 1 will be those that are constant in each subset:
%
\begin{equation}\label{eq:setconst}
  VU\mathbf{x}=\mathbf{x}
  \qquad\means\qquad
  \mathbf{x} = \sum_\alpha x_\alpha \onev^\alpha.
\end{equation}
%

This allows us to write the lumpability condition \eqref{eq:lump}, and the transition matrix for the lumped process compactly:
%
\begin{equation}\label{eq:lumpcompact}
  VU\MM V = \MM V,
  \qquad
  \widehat{\MM} = U\MM V.
\end{equation}
%
By induction, one can show that similar relations hold for all powers:
%
\begin{equation}\label{eq:lumppower}
  VU\MM^nV = \MM^nV,
  \qquad
  \widehat{\MM}^n = U\MM^n V,
\end{equation}
%
and, via the Taylor series, for the exponential as well:
%
\begin{equation}\label{eq:lumpexp}
  VU \e^{t\MM} V = \e^{t\MM}V,
  \qquad
  \e^{t\widehat{\MM}} = U\e^{t\MM} V.
\end{equation}
%
The equilibrium distribution of the lumped process is given by
%
\begin{equation}\label{eq:lumpeq}
  \widehat{\mathbf{p}}^\infty = \eq V.
\end{equation}
%

\section{Signal-to-Noise ratio (SNR)}\label{sec:SNR}

In this section we will look at the signal-to-noise curve, and put an upper bound on its initial value.
We need only consider ergodic Markov chains.
Transient states would be unoccupied in equilibrium and would not be accessed by the signal creation process, therefore they could be removed from the analysis.
Absorbing chains are degenerate cases: they have zero initial signal but infinite decay times, so they can only be approached as the limit of a sequence of ergodic chains.

\subsection{Framework}\label{sec:framework}

The individual potentiation/depression events will be described by \emph{discrete}-time Markov chains:
%
\begin{equation}\label{eq:MWdef}
  \M\potdep \equiv \I + \W\potdep,
  \qquad
  \M\potdep\onev = \onev,
  \qquad
  \M\potdep_{ij} \in [0,1].
\end{equation}
%
The initial signal creation event occurs at time $t=0$, but all subsequent potentiation/depression events occur at random times according to Poisson processes with rates $rf\potdep$, where $f\pot +f\dep =1$ are the fraction of plasticity events that are potentiating/depressing respectively.
This means that the ``forgetting'' process will be described by the \emph{continuous}-time Markov chain:
%
\begin{equation}\label{eq:forgetting}
  \MM = r \frg \equiv r\prn{f\pot\W\pot + f\dep\W\dep}.
\end{equation}
%
We only require that this Markov chain is ergodic.
The Markov chains described by $\M\potdep $ need not be.

We assume that the probability distribution starts in the equilibrium distribution \eqref{eq:equilibrium}.
During the initial signal creation, a fraction $f\pot$ will change to $\eq\M\pot$ and a fraction $f\dep$ will change to $\eq\M\dep$.
After this, probabilities will evolve according to \eqref{eq:rowsum}.

\subsection{SNR curve}\label{sec:SNRcurve}

As discussed in the main text, the signal-to-noise ratio is given by
%
\begin{equation}\label{eq:SNRdef}
  \snr(t) = \frac{\av{\wv_\text{ideal}\cdt\wv(t)} - \av{\wv_\text{ideal}\cdt\wv(\infty)}}
     {\sqrt{\var(\wv_\text{ideal}\cdt\wv(\infty))}}.
\end{equation}
%
First, let's look at the denominator, remembering that the states and plasticity events of each synapse are independent and identically distributed:
%
\begin{equation}\label{eq:noise}
\begin{aligned}
  \var(\wv_\text{ideal}\cdt\wv(\infty))
    &= \sum_{\alpha\beta} \av{\wv^\alpha_\text{ideal}\wv^\alpha(\infty) \wv^\beta_\text{ideal}\wv^\beta(\infty)}
    - \prn{\sum_\alpha \av{\wv^\alpha_\text{ideal}\wv^\alpha(\infty)}}^2 \\
    &= \sum_{\alpha} \av{(\wv^\alpha_\text{ideal})^2(\wv^\alpha(\infty))^2}
    + \sum_{\alpha\neq\beta} \av{\wv^\alpha_\text{ideal}\wv^\alpha(\infty)}\av{\smash{\wv^\beta}_\text{ideal}\smash{\wv^\beta}(\infty)}
    \\&\phantom{= \sum_{\alpha}  \av{(\wv^\alpha_\text{ideal})^2(\wv^\alpha(\infty))^2}}
    - \prn{\sum_\alpha \av{\wv^\alpha_\text{ideal}\wv^\alpha(\infty)}}^2  \\
    &= N\av{1}
    + N(N-1)\av{\wv^1_\text{ideal}\wv^1(\infty)}^2
    - N^2\av{\wv^1_\text{ideal}\wv^1(\infty)}^2 \\
    &= N(1-\av{\wv^1_\text{ideal}\wv^1(\infty)}^2),
\end{aligned}
\end{equation}
%
where we used $\wv^\alpha=\pm1$.
